Ion pairing controls rheological properties of ``processionary'' polyelectrolyte hydrogels

### 2016-08-08 PALSE/talk_Kyoto.tex
fixes from Sebastien comments: coherent colors between strain sweeps and stretch
funny pics for intro on interdisciplinary
Help the flow Counterion condensation -> Behaviour in water extension parameter -> compaction parameter Extension of the electrostatic blobs -> Measure reduced temperature Reorganise conclusion order

\usepackage{pgfplots}
\usepackage{pgfplotstable}
\pgfplotsset{compat=1.9}
\usepgfplotslibrary{groupplots}
\usepgfplotslibrary{external}
\makeatletter
\newcommand*{\overlaynumber}{\number\beamer@slideinframe}
\tikzset{
  beamer externalizing/.style={%
    execute at end picture={%
      \tikzifexternalizing{%
        \ifbeamer@anotherslide
        \pgfexternalstorecommand{\string\global\string\beamer@anotherslidetrue}%
        \fi
      }{}%
    }%
  },
  external/optimize=false
}
\let\orig@tikzsetnextfilename=\tikzsetnextfilename
\renewcommand\tikzsetnextfilename[1]{\orig@tikzsetnextfilename{#1-\overlaynumber}}
\makeatother

@@

@@

\begin{frame}{Physical polyelectrolytes hydrogel}
	\tikzsetnextfilename{polyelectrolytes}%
\input{fig_presentation/polyelectrolytes.pgf}
\begin{columns}
\column{0.5\textwidth}
\structure{Reversible gel}
\begin{itemize}
	\item in deionised, neutral water
	\item yield stress fuild
\end{itemize}
\structure{Solution} if
\begin{itemize}
	\item ionic strength, pH
	\item non ionic head
\end{itemize}
$\Rightarrow$ Head-to-body ionic bonds

\column{0.5\textwidth}
\hfill Herschel-Buckley law\\
\hfill $\sigma =\sigma_c + A \dot{\gamma}^n$
\tikzsetnextfilename{flowcurves}%
\input{fig_presentation/flowcurves.pgf}
\end{columns}
\vspace{-\baselineskip}\textit{\scriptsize Srour et al. Macromol. Rapid Comm., 36, 55 (2015)}
\end{frame}

\begin{frame}{Linear rheology}
	\begin{columns}
	\column{0.5\textwidth}
	\tikzsetnextfilename{linear_rheology_PImBr}%
	\input{fig_presentation/linear_rheology_PImBr.pgf}
	\[ G = \frac{c}{N}k_\mathrm{B}T \]
	\begin{description}[$N$]
	\item[$c$] monomer concentration
	\item[$N$] \# monomers between CL
	\end{description}
	\column{0.45\textwidth}
	\pause
	\begin{itemize}
	\item $N \approx 800 N_0$ !
	\item 800 chains between CL
	\end{itemize}
	\begin{flushright}\includegraphics[height=14\baselineskip]{presentation/Thaumetopea_pityocampa_01.jpg}\\
	\vspace{-1.5em}\colorbox{lightgray}{\scriptsize Wikimedia commons}
	\end{flushright}
	\end{columns}
\end{frame}

\begin{frame}{Tweak repeated ions + counterions}
	\vspace{1.5\baselineskip}
	\tikzsetnextfilename{postfunc}%
	\input{fig_presentation/postfunc.pgf}

	\bigskip
	\tikzsetnextfilename{postfunc_chem}%
	\input{fig_presentation/postfunc_chem.pgf}
\end{frame}

\begin{frame}<1>[label=condensationranking]{Counterion condensation}
	\tikzsetnextfilename{solubility}%
	\input{fig_presentation/solubility.pgf}
	\begin{itemize}
		\item Too much condensation prevents dissolution
		\begin{itemize}
			\item water poor solvent for polymer backbone
			\item polyelectrolytes solubilized by electrostatic repulsion
		\end{itemize}
		\item<2-> Direct Hoffmeister series $\ce{I-}\alert{<}\ce{Br-}\alert{<}\ce{Cl-}\alert{<}\ce{F-}$\\
	\hfill\textit{\scriptsize Hofmeister 1888; Zhang Annu Rev Phys Chem 2010; Schwierz Langmuir 2010}
		\item<3-> \ce{\pi+-\pi+} interactions trap counterions\raisebox{-0.5em}{%
			\begin{tikzpicture}[every node/.style={inner ysep=0,font=\tiny\setatomsep{1.5em}}]
			\node[left=0.5em] (Pyr){\chemfig{[0]*5(--N(-)---)}};
			\node[right=0.5em] (Im){\chemfig{[2]*5(-N(-)-=N-=)}};
			\node[font=\normalsize]{$\alert{<}$};
			\end{tikzpicture}}\\
			\hfill\textit{\scriptsize
			Geronimo PCCP 2011;
			Matthews PCCP 2014;
			Gao J Mol Graph Model 2015}
	\end{itemize}
\end{frame}

@@

	\begin{itemize}
		\item extension parameter $\alert{B} = \frac{\text{extended length}}{\text{diameter}}$ of electrostatic blob
		\smallskip
		\item scales $\geq D$ depend only on $B \sim A^{4/3} \tau$
	\end{itemize}

@@

	\vspace{-0.5\baselineskip}
	\begin{overprint}%
	\onslide<-2>%
	\begin{itemize}
		\item narrow linear regime $\gamma_0\ll 1$
		\item<2> extension parameter increases with condensation
	\end{itemize}
	\onslide<3->%
	\begin{itemize}
		\item bonding energy increases with $B$
		\item<4> measure of counterion condensation $A$
%		\begin{description}
%			\item[\ce{PPyr+I-}] $10$ uncondensed counterions per polymer
%			\item[\ce{PIm+Br-}] a single dissociated counterion every $9$ polymers
%		\end{description}
	\end{itemize}
	\end{overprint}
\end{frame}

@@

\begin{frame}{Extension of the electrostatic blobs}
\vspace{\baselineskip}

\tikzsetnextfilename{stretch}%
	\tikzsetnextfilename{stretch}%
	\input{fig_presentation/stretch.pgf}
	\[E_\mathrm{stretch} = \text{area increment}(\gamma,B) \times \text{surface tension}(\tau,B)\]

	\begin{itemize}
		\item fluidisation = rupture $\Rightarrow E_\mathrm{stretch}(\gamma_c) = E_\text{head-body}$
		\begin{itemize}
			\item \structure{$\tau(E_\text{head-body}, B, \gamma_c)$}
		\end{itemize}
		\item ionic bond energy depends on local $\epsilon_r$%\\\hfill\textit{\scriptsize Rehm Chem Soc Rev. 2010}
		\begin{itemize}
			\item in water $E_\text{head-body}\approx 2k_\mathrm{B}T$
			\item large $B\Rightarrow$ low $\epsilon_r$ $\Rightarrow$ high $E_\text{head-body}$
		\end{itemize}
		\item smallest $B$ is \ce{PPyr+I-} $\Rightarrow \textcolor{ilmcolor}{\tau = 0.4\pm0.07}, \Theta\approx\SI{220}{\celsius}$
	\end{itemize}
\end{frame}

@@

>>>>

\usepackage{pgfplots}
\usepackage{pgfplotstable}
\pgfplotsset{compat=1.9}
\usepgfplotslibrary{fillbetween}
\usepgfplotslibrary{groupplots}
\usepgfplotslibrary{external}
\makeatletter
\newcommand*{\overlaynumber}{\number\beamer@slideinframe}
\tikzset{
  beamer externalizing/.style={%
    execute at end picture={%
      \tikzifexternalizing{%
        \ifbeamer@anotherslide
        \pgfexternalstorecommand{\string\global\string\beamer@anotherslidetrue}%
        \fi
      }{}%
    }%
  },
  external/optimize=false
}
\let\orig@tikzsetnextfilename=\tikzsetnextfilename
\renewcommand\tikzsetnextfilename[1]{\orig@tikzsetnextfilename{#1-\overlaynumber}}
\makeatother

@@

\begin{frame}{What chemists/physicists think I do}
	\begin{columns}[t]
	\column{0.5\textwidth}
	\begin{flushleft}\structure{chemistry}

	``This counterion is more chaotropic''

	\vspace{\baselineskip}
	\includegraphics[width=\textwidth]{presentation/sorciere}
	\end{flushleft}
	\column{0.5\textwidth}
	\begin{flushright}\structure{rheology}

	``We impose a constant shear stress''
	\end{flushright}
	\hspace{0.25\textwidth}\includegraphics[height=10\baselineskip]{presentation/rascar}
	\end{columns}
\end{frame}

\begin{frame}{Physical chemistry positive feedback}
\tikzsetnextfilename{chem_rheol}%
	\input{fig_presentation/chem_rheol.pgf}
\end{frame}

@@

\begin{frame}{Physical polyelectrolytes hydrogel}
	\tikzsetnextfilename{polyelectrolytes}%
\input{fig_presentation/polyelectrolytes.pgf}
\begin{columns}
\column{0.5\textwidth}
\structure{Reversible gel}
\begin{itemize}
	\item in deionised, neutral water
	\item yield stress fluid
\end{itemize}
\structure{Solution} if
\begin{itemize}
	\item ionic strength, pH
	\item non ionic head
\end{itemize}
$\Rightarrow$ Head-to-body ionic bonds

\column{0.5\textwidth}
\hfill Herschel-Bulkley law\\
\hfill $\sigma =\sigma_c + A \dot{\gamma}^n$
\tikzsetnextfilename{flowcurves}%
\input{fig_presentation/flowcurves.pgf}
\end{columns}
\vspace{-\baselineskip}\textit{\scriptsize Srour et al. Macromol. Rapid Comm., 36, 55 (2015)}
\end{frame}

\begin{frame}{Linear rheology: A physicist problem}
	\begin{columns}
	\column{0.5\textwidth}
	\tikzsetnextfilename{linear_rheology_PImBr}%
	\input{fig_presentation/linear_rheology_PImBr.pgf}
	\[ G = \frac{c}{N}k_\mathrm{B}T \]
	\begin{description}[$N$]
	\item[$c$] monomer concentration
	\item[$N$] \# monomers between CL
	\end{description}
	\column{0.45\textwidth}
	\pause
	\begin{itemize}
	\item $N \approx 800 N_0$ !
	\item 800 chains between CL
	\end{itemize}
	\begin{flushright}\includegraphics[height=14\baselineskip]{presentation/Thaumetopea_pityocampa_01.jpg}\\
	\vspace{-1.5em}\colorbox{lightgray}{\scriptsize Wikimedia commons}
	\end{flushright}
	\end{columns}
\end{frame}

\begin{frame}{Tweaking repeated ions + counterions}
	\vspace{1.5\baselineskip}
	\tikzsetnextfilename{postfunc}%
	\input{fig_presentation/postfunc.pgf}

	\bigskip
	\tikzsetnextfilename{postfunc_chem}%
	\input{fig_presentation/postfunc_chem.pgf}

	\vspace{-3\baselineskip}\hfill \begin{minipage}{0.35\textwidth}``Can you put that in the rheometer?''\end{minipage}
\end{frame}

\begin{frame}<1>[label=condensationranking]{Behaviour in water}
	\tikzsetnextfilename{solubility}%
	\input{fig_presentation/solubility.pgf}
	\begin{itemize}
		\item Too much condensation prevents dissolution
		\begin{itemize}
			\item water poor solvent for polymer backbone
			\item polyelectrolytes solubilized by electrostatic repulsion
		\end{itemize}
		\item<2-> Direct Hoffmeister series $\ce{I-}\alert{<}\ce{Br-}\alert{<}\ce{Cl-}\alert{<}\ce{F-}$\\
	\hfill\textit{\scriptsize Hofmeister 1888; Zhang Annu Rev Phys Chem 2010; Schwierz Langmuir 2010}
		\item<3-> \ce{\pi+-\pi+} interactions trap counterions\raisebox{-0.5em}{%
			\begin{tikzpicture}[every node/.style={inner ysep=0,font=\tiny\setatomsep{1.5em}}]
			\node[left=0.5em] (Pyr){\chemfig{[0]*5(--N(-)---)}};
			\node[right=0.5em] (Im){\chemfig{[2]*5(-N(-)-=N-=)}};
			\node[font=\normalsize]{$\alert{<}$};
			\end{tikzpicture}}\\
			\hfill\textit{\scriptsize
			Geronimo PCCP 2011;
			Matthews PCCP 2014;
			Gao J Mol Graph Model 2015}
		\item<3-> 5 gels ranked by counterion condensation
	\end{itemize}
\end{frame}

@@

	\begin{itemize}
		\item compaction parameter $\alert{B} = \frac{\text{extended length}}{\text{diameter}}$ of electrostatic blob
		\smallskip
		\item scales $\geq D$ depend only on $B \sim A^{4/3} \tau$
	\end{itemize}

@@

	\vspace{-0.5\baselineskip}
	\begin{overprint}%
	\onslide<-2>%
	\begin{itemize}
		\item narrow linear regime $\gamma_0\ll 1$
		\item<2> compaction parameter increases with condensation
	\end{itemize}
	\onslide<3>%
	\begin{itemize}
		\item bonding energy increases with compaction (condensation)
		\item allows larger strains
	\end{itemize}
	\onslide<4>%
	\begin{itemize}
		\item $A \sim (B/\tau)^{3/4}$
		\item measure of counterion condensation $A$
	\end{itemize}
	\end{overprint}
\end{frame}

@@

\begin{frame}{Measure reduced temperature}
\vspace{\baselineskip}
\begin{itemize}
		\item fluidisation = rupture: stretching breaks head-to-body
\end{itemize}
\tikzsetnextfilename{stretch}%
	\tikzsetnextfilename{stretch}%
	\input{fig_presentation/stretch.pgf}
	\[E_\mathrm{stretch} = \text{area increment}(\gamma,B) \times \text{surface tension}(\textcolor{ilmcolor}{\tau},B)\]

	\begin{itemize}
		%\item \structure{$\tau(E_\text{head-body}, B, \gamma_c)$}
		\item for \ce{PPyr+I-}
		\begin{itemize}
			\item little compaction $B\approx 2.4$
			\item ionic bond in water: $E_\text{head-body}\approx 2k_\mathrm{B}T$
			\item $\textcolor{ilmcolor}{\tau = 0.4\pm0.07}\qquad \Theta\approx\SI{220}{\celsius}$
		\end{itemize}
		\item ionic bond energy depends on local $\epsilon_r$%\\\hfill\textit{\scriptsize Rehm Chem Soc Rev. 2010}
		\begin{itemize}
			\item large $B\Rightarrow$ large electrostatic blobs with low $\epsilon_r$
			\item stronger head-to-body bonds
		\end{itemize}
	\end{itemize}
\end{frame}

@@

\begin{frame}{Conclusion}
	\begin{itemize}
		\item post-functionalisation allows to change $A\times 100$
		\begin{itemize}
			\item \ce{PPyr+I-}: 10 free counterions per polymer
			\item \ce{PIm+Br-}: 1 free counterion every 9 polymers
		\end{itemize}
		\item mechanical properties $\times 1000$
		\begin{itemize}
			\item \ce{PPyr+I-}: $G\approx\SI{10}{\kilo\pascal}$, fluidizes at $\gamma_c\approx 2\%$
			\item \ce{PIm+Br-}: $G\approx\SI{10}{\pascal}$, fluidizes at $\gamma_c\approx 800\%$
		\end{itemize}
		\item measure of microscopic parameters from rheology
		\begin{itemize}
			\item counterion condensation
			\item head-to-body bonding energy
		\end{itemize}
		\item highly tunable hydrogels from short, linear polymer chains
		\begin{itemize}
			\item no additive
			\item yield stress fluids
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Processionary phase diagram}
	\begin{columns}
	\column{0.65\textwidth}
	\vspace{\baselineskip}
	\tikzsetnextfilename{phase_diag}%
	\input{fig_presentation/phase_diag.pgf}
	\column{0.35\textwidth}
	\begin{flushright}\includegraphics[height=14\baselineskip]{presentation/Thaumetopea_pityocampa_01.jpg}\\
	\vspace{-1.5em}\colorbox{lightgray}{\scriptsize Wikimedia commons}
	\end{flushright}
	\end{columns}
\end{frame}
